\bf Model Visualization Techniques for a Social Network Model

### 2018-05-14 paper-child.Rnw
getting there :pensive:

%These three uses of the term ``model"
\st{The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can }view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result. \st{Since we have already decided on our model family, SAOMs, }we now shift our focus to the fitted model and the model form. Specifically, we want to learn more about how the model form we choose affects the fitted model by using our example data sets and our visualization toolbox. We begin by introducing the five models that we fit to our example data. \st{Next we use the five models to guide our visual explorations of SAOMs. We first use novel tools and ideas to view a SAOM in the data space of a dynamic network. We then explore collections of the same models fit many times to the example data to see how the simulation processes in \texttt{RSiena} affect the model fits. Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this "hidden data" mechanism works and how it results in a fitted model. }

@@

We first consider the 16 actor subset of the teenage friends and lifestyle data available on the \texttt{RSiena} website \citep{RSiena}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align*}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
\end{align*}

@@

where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$. These statistics are known as the number of jumping transitive triplets and the number of doubly achieved distances two effect, respectively. The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors. The covariate groups are determined by the student's drinking behavior, the values of which are numeric and mean-centered. The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$. The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.

In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison. \st{For fitting M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable. In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.} %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

%In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}. The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$. The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
\st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.

@@

The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space. In \citeauthor{modelvis}, they \st{happen to} define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for network data? For dynamic social networks, there are a few different data ``spaces":
\begin{enumerate}
\item the actors and their corresponding covariates,
\item the edges and their correpsonding variables that describe the relationships between the nodes, and
\item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
\end{enumerate}
These three data pieces can be visualized together in various ways. The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

\st{Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing. %The SIENA software includes a method for modelling covariate change in the network
Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time. In this instance, we would want to view predictions over time. A SAOM can also model behavior change over time, taking both the node and edge information into account. In this case, a plot of predicted covariate values over time would put the model into the time and node data space. Most likely, however, is that we would want to view all of the data spaces simultaneously.}

>>>>

%These three uses of the term ``model"
The model family, the model form, and the fitted model can each be visualized according to the three principals of model visualization: we can view the model in the data space, visualize collections of models, and explore the process of fitting the model, not just the end result. Since we have already decided on our model family, SAOMs, we shift our focus to the fitted model and the model form. Specifically, we want to learn more about how the \emph{model form} we choose affects the \emph{fitted model} by using our example data sets and our visualization toolbox. We begin by introducing the five models that we fit to our example data. We then use the five models to guide our visual explorations of SAOMs. We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network. We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits. Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.

@@

We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figures \ref{fig:adjmatviz, fig:alldataviz}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align*}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
\end{align*}

@@

where $s_{i3} = \sum_{j \neq h} x_{ij}x_{ih}x_{hj} \mathbb{I}(z_i = z_h \neq z_j)$, and $s_{i4} =  |\{j : x_{ij} = 0, \sum\limits_h x_{ih}x_{hj} \geq 2\}|$. These statistics are known as the \emph{number of jumping transitive triplets} and the \emph{number of doubly achieved distances two effect}, respectively. The first statistic emphasizes triad relationships that are formed between actors from different covariate groups, while the other emphasizes indirect ties between actors. The covariate groups are determined by the student's drinking behavior, the values of which are converted to numeric then mean-centered. The four values in our data are $\{\text{never, 1-2 times per year, once per month, once per week}\}$, which after being converted to numeric and mean-centered become $\{-0.8125,-1.8125,0.1875,1.1875\}$. The two additional effects are visually represented and further described in Figure~\ref{fig:jtt} and Figure~\ref{fig:dad}, respectively.

In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison. To fit M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable. In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}.
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

@@

%In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}. The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$. The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties.
Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. By graphically exploring these models, we aim to understand their effects, their interpretations, and the model fitting process better.

@@

The first way we hope to better understand stochastic actor-oriented models is by viewing the model(s) in the data space. In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for network data? The multi-level data structure of networks makes answering this question difficult.
% %For dynamic social networks, there are a few different data ``spaces":
% \begin{enumerate}
% \item the actors and their corresponding covariates,
% \item the edges and their correpsonding variables that describe the relationships between the nodes, and
% \item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
% \end{enumerate}
The node, edge, and time data can be visualized together in various ways. %The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing. %The SIENA software includes a method for modelling covariate change in the network
Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time. In this instance, we would want to view predictions over time. A SAOM can also model behavior change over time, taking both the node and edge information into account. In this case, a plot of predicted covariate values over time would put the model into the time and node data space. Most likely, however, is that we would want to view all of the data spaces simultaneously.
